---
title: "Happiness on a map"
author: Raphaël Mauron
date: today
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

Here, we can use an interactive map to visualize the input data and some of the results.

## Load, explore and prepare the data

First, we need to load and make sure of what we have as data.

```{r}
#| label: setup
#| message: false

# Not cached: on a cache hit knitr restores the chunk's *objects* but does not
# re-attach its packages, so every later chunk loses str_remove(), glimpse(),
# pivot_wider() and friends. Attaching them takes a second anyway.
library(dplyr)
library(stringr)
library(ggplot2)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale
library(readr)
library(giscoR)
library(sf)
library(leaflet)
library(patchwork)
library(DT)
library(tidyverse)
```


```{r}
#| label: load-data-wide
#| message: false
data_wide <- read_csv("./data_wide_before_croatia.csv")
dplyr::glimpse(data_wide)
```


## Add geometry/maps borders

To prepare the plotting into maps, we need some more information, notably the 
borders of the country present.

`giscoR` serves the official Eurostat GISCO boundaries


```{r}
#| label: geometry
#| cache: true

# Cached, and nothing invalidates it automatically: `dependson` only tracks
# chunks that are themselves cached, and load-data-wide is not. If the country
# set in data_wide ever changes, delete analysis_4_cache/ by hand.

borders <- giscoR::gisco_get_countries( #get coordinates of the country present in our data_wide$iso3
  resolution = "20",              # 1:20 million — plenty for a country map
  epsg       = "4326",
  country    = data_wide$iso3
)

borders <- sf::st_make_valid(borders) #check whether the borders are valid for sf

# France, Spain, Portugal, Norway carry overseas/Arctic territory in GISCO,
# so never trust the automatic bounds — set them explicitly.
eu_bbox <- c(xmin = -12, ymin = 34, xmax = 32, ymax = 71)

head(borders)
```

Borders is a tibble with each row representing a single country's borders information.

### Join data_wide and borders

Since for the plotting we want each country to appear only exactly once, we have to
select 

```{r}
#| label: map-data

# The slice the maps are fixed at. "Doctorate"/"Master's" x "All ages" are the
# only combinations that carry all six measures (see the coverage note above).
map_education <- "Master's"
map_age_group <- "All ages"

# `borders` stays geometry-only: one row per country, no measures attached.
# data_wide has 11 rows per country (education x age band), so the join has to
# happen against one slice at a time — see the map chunk.
eu_sf <- borders |>
  left_join(filter(data_wide, education == map_education, age_group == map_age_group),
            by = c("ISO3_CODE" = "iso3"))

stopifnot(nrow(eu_sf) == nrow(borders))   # one row, one geometry, one country

metrics <- tribble(
  ~col,                     ~label,                     ~palette,
  "pct_happy",              "Happy (%)",                "YlGnBu",
  "pct_satisfied_finances", "Satisfied, finances (%)",  "YlGnBu",
  "pct_satisfied_leisure",  "Satisfied, leisure (%)",   "YlGnBu",
  "pct_permanent_contract", "Permanent contract (%)",   "YlGnBu",
  "earnings_gross_yearly",  "Gross yearly earnings",    "YlOrBr",
  "hours_worked",           "Hours worked per year",    "YlOrBr"
)

# Fail loud if a slice ever stops carrying one of the layers.
stopifnot(all(metrics$col %in% names(eu_sf)),
          !sapply(metrics$col, \(c) all(is.na(eu_sf[[c]]))))
```



### All six metrics

Show all the quantitative metrics side by side.

```{r}
#| label: metric-grid
#| fig-width: 12
#| fig-height: 8
map_panel <- function(col, label, palette) {
  ggplot(eu_sf) +
    geom_sf(aes(fill = .data[[col]]), colour = "white", linewidth = 0.15) +
    coord_sf(xlim = eu_bbox[c("xmin", "xmax")],
             ylim = eu_bbox[c("ymin", "ymax")], expand = FALSE) +
    # direction = 1 because distiller defaults to -1, which would paint low
    # values dark and silently invert every panel against the leaflet map.
    scale_fill_distiller(palette = palette, direction = 1, na.value = "grey90",
                         labels = scales::label_number(big.mark = " ")) +
    labs(title = label, fill = NULL) +
    theme_void(base_size = 9) +
    theme(
      plot.title        = element_text(size = 10, hjust = 0, margin = margin(b = 4)),
      # without the bottom margin the row-1 colourbars land on the row-2 titles
      # and the row-2 colourbars get clipped by the figure edge.
      plot.margin       = margin(4, 6, 14, 6),
      legend.position   = "bottom",
      legend.margin     = margin(t = 6, b = 0),
      legend.text       = element_text(size = 7),
      legend.key.height = unit(3, "pt"),
      legend.key.width  = unit(28, "pt")
    )
}

wrap_plots(pmap(metrics, map_panel), nrow = 2, ncol = 3) +
  plot_annotation(title = paste(map_education, "-", map_age_group))
```

  
## Correct the income values of Croatia:

Croatia reports earnings in kuna while every other country is on the survey's
harmonised unit. Mean gross yearly earnings across all EARNY variables:
HRV 322 743, against 122 411 for Luxembourg, the next highest -- a 2.6x gap
with no counterpart in any other indicator. Dividing by the irrevocable

Dividing by the irrevocable
HRK/EUR rate fixed at Croatia's euro entry (1 Jan 2023) lands Croatia at
42 835, between Estonia and Poland, which matches its position elsewhere.


```{r}
HRK_PER_EUR <- 7.53450

# Before/after, for the write-up. Run this *before* the mutate below.
data_wide |>
  filter(iso3 == "HRV", !is.na(earnings_gross_yearly)) |>
  transmute(education, age_group,
            as_reported = round(earnings_gross_yearly),
            in_euro     = round(earnings_gross_yearly / HRK_PER_EUR))

data_wide <- data_wide |>
  dplyr::mutate(earnings_gross_yearly = if_else(iso3 == "HRV",
                                         earnings_gross_yearly / HRK_PER_EUR,
                                         earnings_gross_yearly))

```


## The slice both maps draw

Prepare the map data with the corrected data.

```{r}
#| label: map-data2

# The slice the maps are fixed at. "Doctorate"/"Master's" x "All ages" are the
# only combinations that carry all six measures (see the coverage note above).
map_education <- "Master's"
map_age_group <- "All ages"

eu_sf <- borders |>
  left_join(filter(data_wide, education == map_education, age_group == map_age_group),
            by = c("ISO3_CODE" = "iso3"))

stopifnot(nrow(eu_sf) == nrow(borders))   # one row, one geometry, one country
```


## All six metrics, Croatia corrected

```{r}
#| label: metric-grid2
#| fig-width: 12
#| fig-height: 8
map_panel <- function(col, label, palette) {
  ggplot(eu_sf) +
    geom_sf(aes(fill = .data[[col]]), colour = "white", linewidth = 0.15) +
    coord_sf(xlim = eu_bbox[c("xmin", "xmax")],
             ylim = eu_bbox[c("ymin", "ymax")], expand = FALSE) +
    # direction = 1 because distiller defaults to -1, which would paint low
    # values dark and silently invert every panel against the leaflet map.
    scale_fill_distiller(palette = palette, direction = 1, na.value = "grey90",
                         labels = scales::label_number(big.mark = " ")) +
    labs(title = label, fill = NULL) +
    theme_void(base_size = 9) +
    theme(
      plot.title        = element_text(size = 10, hjust = 0, margin = margin(b = 4)),
      # without the bottom margin the row-1 colourbars land on the row-2 titles
      # and the row-2 colourbars get clipped by the figure edge.
      plot.margin       = margin(4, 6, 14, 6),
      legend.position   = "bottom",
      legend.margin     = margin(t = 6, b = 0),
      legend.text       = element_text(size = 7),
      legend.key.height = unit(3, "pt"),
      legend.key.width  = unit(28, "pt")
    )
}

wrap_plots(pmap(metrics, map_panel), nrow = 2, ncol = 3) +
  plot_annotation(title = paste(map_education, "-", map_age_group))
```


## Master's over Doctorate: the same six metrics

The grid above fixes education, so it cannot show what the education level does to
each measure. Here the six metrics are the columns and the two education levels
are the rows: read down a column to compare Master's against Doctorate, across a
row to compare measures within one level.

```{r}
#| label: metric-grid-education
#| fig-width: 14
#| fig-height: 8
#| column: page

edu_slice <- function(edu) {
  s <- borders |>
    left_join(filter(data_wide, education == edu, age_group == map_age_group),
              by = c("ISO3_CODE" = "iso3"))
  stopifnot(nrow(s) == nrow(borders))   # one row, one geometry, one country
  s
}

sf_mas <- edu_slice("Master's")
sf_phd <- edu_slice("Doctorate")

# Fail if a metric ever goes empty in both slices: `range()`
stopifnot(all(metrics$col %in% names(sf_phd)),
          !sapply(metrics$col, \(c) all(is.na(c(sf_mas[[c]], sf_phd[[c]])))))

# One gradient per metric. All six are ColorBrewer sequential ramps, so each is
# still colour-blind safe on its own. Purples and PuRd are the closest pair of
# the six, which is why they are not neighbouring columns; Greys is left out
# because grey90 below already means "no data". Local to this figure --
# `metrics$palette` still drives the 6-panel grid and the leaflet map.
edu_palettes <- c(
  pct_happy              = "PuRd",
  pct_satisfied_finances = "Blues",
  pct_satisfied_leisure  = "Greens",
  pct_permanent_contract = "Purples",
  earnings_gross_yearly  = "YlOrBr",
  hours_worked           = "OrRd"
)
stopifnot(setequal(names(edu_palettes), metrics$col))

# Per metric, not per panel: both rows of a column read off the same scale.
edu_limits <- setNames(
  lapply(metrics$col, \(col) range(c(sf_mas[[col]], sf_phd[[col]]), na.rm = TRUE)),
  metrics$col
)

# Map function
map_panel_edu <- function(sf_slice, col, palette, limits, title, subtitle) {
  ggplot(sf_slice) +
    geom_sf(aes(fill = .data[[col]]), colour = "white", linewidth = 0.15) +
    coord_sf(xlim = eu_bbox[c("xmin", "xmax")],
             ylim = eu_bbox[c("ymin", "ymax")], expand = FALSE) +
    scale_fill_distiller(palette = palette, direction = 1, limits = limits,
                         na.value = "grey90",
                         labels = scales::label_number(big.mark = " ")) +
    labs(title = title, subtitle = subtitle, fill = NULL) +
    theme_void(base_size = 8) +
    theme(
      plot.title        = element_text(size = 8, hjust = 0, margin = margin(b = 1)),
      plot.subtitle     = element_text(size = 7, hjust = 0, colour = "grey30",
                                       margin = margin(b = 4)),
      plot.margin       = margin(4, 4, 12, 4),
      legend.position   = "bottom",
      legend.text       = element_text(size = 6),
      legend.key.height = unit(3, "pt"),
      legend.key.width  = unit(20, "pt")
    )
}


edu_row <- function(sf_slice, edu) {
  pmap(metrics, function(col, label, ...)    # gradient comes from edu_palettes
    map_panel_edu(sf_slice, col, edu_palettes[[col]], edu_limits[[col]],
                  label, edu))
}

wrap_plots(c(edu_row(sf_mas, "Master's"), edu_row(sf_phd, "Doctorate")),
           nrow = 2, ncol = 6) +
  plot_annotation(title = paste("Master's over Doctorate -", map_age_group))
```


## We can explore it one by one

Using leaflet, we can switch to the different layers easily!

The radio buttons can only carry one dimension, and one geometry can only carry
one row, so education and age band are fixed and the six measures are the
toggle. 

Coverage forces that choice: the wellbeing measures are reported only at
`All ages`, the labour measures only for Master's and Doctorate. `Master's` and
`Doctorate` at `All ages` are therefore the only two slices with all six.

In that case we fixed the age to `All ages` and education to `Master's`.


```{r}
#| label: leaflet-map2

add_metric_layer <- function(map, col, label, palette) {
  v   <- eu_sf[[col]]
  pal <- colorNumeric(palette, domain = v, na.color = "#e8e8e8")
  # 4-6 countries are missing per measure; say so rather than printing "NA".
  txt <- ifelse(is.na(v), "no data",
                formatC(v, format = "f", digits = 1, big.mark = " "))

  map |>
    addPolygons(
      data         = eu_sf,
      group        = label,
      fillColor    = ~ pal(v),
      fillOpacity  = 0.8,
      color        = "white",
      weight       = 1,
      label        = ~ sprintf("%s: %s", NAME_ENGL, txt) |> lapply(htmltools::HTML),
      highlightOptions = highlightOptions(weight = 3, color = "#333", bringToFront = TRUE)
    ) |>
    addLegend(
      position = "bottomright", pal = pal, values = v,
      title = label, group = label, opacity = 0.9
    )
}

m <- leaflet(options = leafletOptions(minZoom = 3)) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>
  fitBounds(eu_bbox[["xmin"]], eu_bbox[["ymin"]], eu_bbox[["xmax"]], eu_bbox[["ymax"]])

m <- reduce(
  seq_len(nrow(metrics)),
  \(map, i) add_metric_layer(map, metrics$col[i], metrics$label[i], metrics$palette[i]),
  .init = m
)

m |>
  addLayersControl(
    baseGroups    = metrics$label,           # radio buttons: one metric at a time
    options       = layersControlOptions(collapsed = FALSE)
  ) |>
  # leaflet ties legend visibility to overlayadd/overlayremove/groupadd/
  # groupremove and has no baselayerchange handler, so with baseGroups every
  # legend is added at init and never removed -- all six stack up at once.
  # Translate the event the radio buttons do fire into the ones it listens for.
  htmlwidgets::onRender("
function(el, x) {
  var map = this;
  var groups = x.calls
    .filter(function(c) { return c.method === 'addLegend'; })
    .map(function(c) { return c.args[0].group; });

  // addLayersControl already leaves only the first polygon layer visible;
  // do the same for the legends.
  groups.slice(1).forEach(function(g) { map.fire('groupremove', {name: g}); });

  var shown = groups[0];
  map.on('baselayerchange', function(e) {
    if (e.name === shown) return;
    map.fire('groupremove', {name: shown});
    map.fire('groupadd',    {name: e.name});
    shown = e.name;
  });
}")
```


## Two metrics side by side

Do any of these metrics actually correlate?

With the ShinyMap bellow, you can select any two of the metrics, and it will
return there corresponding maps and there correlation scores, for both education
level simultaneously.

The interactivity we had above is not enough to let us do that.

We need our new [`shinyMap.R`](shinyMap.R) kept out of the `.qmd` to avoid rendering
issues.

The app draws the same `data_wide` slices and reuses the leaflet recipe from
`add_metric_layer()` above — `colorNumeric()` fill, `no data` labels for the 4-6
countries missing per measure, the same `CartoDB.PositronNoLabels` tiles and the
same fixed `eu_bbox`. 

What you can do:
1. Select two metrics
2. Visualize the correlation directly
3. Click any country on the maps or on the scatter plots.
4. Deepen your request by selecting age categories.

Each scatter reports its own Pearson and Spearman correlation with the `n` it was
computed from, and a toggle puts both rows of a column on one shared colour scale 
— the `edu_limits` trick from the twelve-map grid, without which the darkest shade
means something different in the Master's map than in the Doctorate one.


```{r}
#| label: launch-shiny
#| eval: false

# From the project root. eval: false above -- this never runs during a render.
# This is the fast path while developing: real R, instant reload.
shiny::runApp("shinyMap.R")
```


### Getting it into this page: R compiled to WebAssembly

The chunk above needs a live R process, which a rendered `.html` does not have.
To make the app work *in this page* the R process has to move into the reader's
browser, and that is what `shinylive` does: it ships webR -- R itself compiled to
WebAssembly -- alongside wasm builds of every package the app loads. No server, no
`shinyapps.io` account, and the result is still a static site.

Every package this app needs is available as a wasm binary from
[`repo.r-wasm.org`](https://repo.r-wasm.org): `shiny`, `bslib`, `dplyr`, `readr`,
`sf`, `leaflet` and `plotly`.

We just needed to save the borders file into a file locally.

```{r}
#| label: save-borders

# 2 decimal places is ~1 km at these latitudes, invisible at 1:20 million, and
# it takes the file from 156 KB to 137 KB. Only the two columns the app uses.
sf::st_write(
  dplyr::select(borders, ISO3_CODE, NAME_ENGL),
  "borders.geojson",
  driver        = "GeoJSON",
  delete_dsn    = TRUE,
  quiet         = TRUE,
  layer_options = "COORDINATE_PRECISION=2"
)
```


So the export lives in [`export_shinylive.R`](export_shinylive.R) and runs as a
`pre-render` step, declared in `_quarto.yml` alongside the `resources:` entry it
has to precede. It downloads webR and a wasm binary for every package the app
loads, so the first build takes minutes; `shinylive` caches them afterwards.
Pre-render scripts are not covered by `freeze: auto` and would otherwise re-export
on every single render, so the script skips the work unless `shinyMap.R` is newer
than the bundle. Delete `shinylive_app/` to force a rebuild after a data change.

```{r}
#| label: check-shinylive-export
#| echo: false

# The runtime, not just the page: index.html on its own renders as a blank frame,
# which is easy to mistake for a broken app rather than a missing export.
if (!dir.exists("shinylive_app/shinylive")) {
  warning("shinylive_app/ is missing or incomplete -- check that the ",
          "pre-render step in _quarto.yml ran; the frame below will be blank.",
          call. = FALSE)
}
```


::: {.column-screen-inset}
```{=html}
<iframe src="shinylive_app/index.html" width="100%" height="1040"
        style="border: 1px solid #dee2e6; border-radius: 4px;">
</iframe>
```
:::


```{r}
#| label: print-time
#| message: false
#| echo: false
sessionInfo()
```