---
title: "Analysis 1"
---

“Does age pay? Earnings, job security and working hours among Master’s and Doctorate holders across countries.”

Analysis of age-related data.
Clean data first.
We restricted the analysis to observations classified by OECD as normal values (OBS_STATUS_1 = "A"). When two normal observations were available for the same country, year and indicator, we used their arithmetic mean and flagged these cases for sensitivity analysis.

```{r}
#| label: setup
#| message: false
#| cache: true

library(tidyverse)
dat <- read_csv("./Scrooge_McDuck/filtered_reICO_data.csv")
head(dat)
colnames(dat)

library(tidyverse)


# Keep only age-specific variables
age_data <- dat %>%
  filter(
    str_detect(
      Indicator,
      "aged 25-34|aged 35-44|aged 45-54|aged 55-64"
    )
  )

# Keep observations classified as "Normal value"
age_normal <- age_data %>%
  filter(OBS_STATUS_1 == "A")

#Coerce duplications into a mean
age_normal_check <- age_normal %>%
  group_by(REF_AREA, `Reference area`, VAR, Indicator) %>%
  summarise(
    n_values = n(),
    value_min = min(OBS_VALUE),
    value_max = max(OBS_VALUE),
    value_mean = mean(OBS_VALUE),
    difference = value_max - value_min,
    pct_difference =
      100 * difference / value_mean,
    .groups = "drop"
  )


#Create tidy table
age_tidy <- age_normal_check %>%
  mutate(
    
    # Degree
    degree = case_when(
      str_starts(VAR, "MAS") ~ "Master",
      str_starts(VAR, "PHD") ~ "Doctorate",
      TRUE ~ NA_character_
    ),
    
    # Age group
    age_group = case_when(
      str_detect(VAR, "25_34") ~ "25-34",
      str_detect(VAR, "35_44") ~ "35-44",
      str_detect(VAR, "45_54") ~ "45-54",
      str_detect(VAR, "55_64") ~ "55-64",
      TRUE ~ NA_character_
    ),
    
    # Outcome
    outcome = case_when(
      str_detect(VAR, "EARNY") ~ "Earnings",
      str_detect(VAR, "_HW")   ~ "Annual hours",
      str_detect(VAR, "PERM")  ~ "Indefinite contract",
      TRUE ~ NA_character_
    ),
    
    # Put age groups in logical order
    age_group = factor(
      age_group,
      levels = c("25-34", "35-44", "45-54", "55-64"),
      ordered = TRUE
    ),
    
    # Put degree categories in logical order
    degree = factor(
      degree,
      levels = c("Master", "Doctorate")
    ),
    
    # Flag cases where more than one normal value was averaged
    duplicate_normal = n_values > 1
  ) %>%
  
  select(
    country_code = REF_AREA,
    country = `Reference area`,
    degree,
    age_group,
    outcome,
    value = value_mean,
    duplicate_normal,
    n_normal_values = n_values,
    value_min,
    value_max,
    difference,
    pct_difference,
    VAR
  ) %>%
  
  arrange(
    country,
    degree,
    outcome,
    age_group
  )

#Check that all variables got converted and no NA values exist
age_tidy %>%
  filter(
    is.na(degree) |
    is.na(age_group) |
    is.na(outcome)
  )

#Inspect coverage
age_tidy %>%
  count(degree, outcome, age_group)

#Check that no duplicate values remain
age_tidy %>%
  count(country, degree, age_group, outcome) %>%
  filter(n > 1)
```

```{r}
#| label: setup
#| message: false
#| cache: true
library(dplyr)
library(ggplot2)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale

```


```{r}
#| label: print-time
#| message: false
cat(format(Sys.time()))
```