---
title: "Analysis 1"
---

“Does age pay? Earnings, job security and working hours among Master’s and Doctorate holders across countries.”

Analysis of age-related data.
Clean data first.
We restricted the analysis to observations classified by OECD as normal values (OBS_STATUS_1 = "A"). When two normal observations were available for the same country, year and indicator, we used their arithmetic mean and flagged these cases for sensitivity analysis.

```{r}
#| label: setup
#| message: false
#| cache: true

library(tidyverse)
dat <- read_csv("./Scrooge_McDuck/filtered_reICO_data.csv")
head(dat)
colnames(dat)

library(tidyverse)


# Keep only age-specific variables
age_data <- dat %>%
  filter(
    str_detect(
      Indicator,
      "aged 25-34|aged 35-44|aged 45-54|aged 55-64"
    )
  )

# Keep observations classified as "Normal value"
age_normal <- age_data %>%
  filter(OBS_STATUS_1 == "A")

#Coerce duplications into a mean
age_normal_check <- age_normal %>%
  group_by(REF_AREA, `Reference area`, VAR, Indicator) %>%
  summarise(
    n_values = n(),
    value_min = min(OBS_VALUE),
    value_max = max(OBS_VALUE),
    value_mean = mean(OBS_VALUE),
    difference = value_max - value_min,
    pct_difference =
      100 * difference / value_mean,
    .groups = "drop"
  )

# Change column names to human readable
age_normal_check %>% 
    rename(Country = "Reference area", )


```

```{r}
#| label: setup
#| message: false
#| cache: true
library(dplyr)
library(ggplot2)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale

```


```{r}
#| label: print-time
#| message: false
cat(format(Sys.time()))
```