---
title: "Analysis 1"
---

“Does age pay? Earnings, job security and working hours among Master’s and Doctorate holders across countries.”

Analysis of age-related data.
Clean data first.
We restricted the analysis to observations classified by OECD as normal values (OBS_STATUS_1 = "A"). When two normal observations were available for the same country, year and indicator, we used their arithmetic mean and flagged these cases for sensitivity analysis.

```{r}
#| label: setup
#| message: false
#| cache: true

library(tidyverse)
dat <- read_csv("./Scrooge_McDuck/filtered_reICO_data.csv")
head(dat)
colnames(dat)

library(tidyverse)


# Keep only age-specific variables
age_data <- dat %>%
  filter(
    str_detect(
      Indicator,
      "aged 25-34|aged 35-44|aged 45-54|aged 55-64"
    )
  )

# Keep observations classified as "Normal value"
age_normal <- age_data %>%
  filter(OBS_STATUS_1 == "A")

#Coerce duplications into a mean
age_normal_check <- age_normal %>%
  group_by(REF_AREA, `Reference area`, VAR, Indicator) %>%
  summarise(
    n_values = n(),
    value_min = min(OBS_VALUE),
    value_max = max(OBS_VALUE),
    value_mean = mean(OBS_VALUE),
    difference = value_max - value_min,
    pct_difference =
      100 * difference / value_mean,
    .groups = "drop"
  )


#Create tidy table
age_tidy <- age_normal_check %>%
  mutate(
    
    # Degree
    degree = case_when(
      str_starts(VAR, "MAS") ~ "Master",
      str_starts(VAR, "PHD") ~ "Doctorate",
      TRUE ~ NA_character_
    ),
    
    # Age group
    age_group = case_when(
      str_detect(VAR, "25_34") ~ "25-34",
      str_detect(VAR, "35_44") ~ "35-44",
      str_detect(VAR, "45_54") ~ "45-54",
      str_detect(VAR, "55_64") ~ "55-64",
      TRUE ~ NA_character_
    ),
    
    # Outcome
    outcome = case_when(
      str_detect(VAR, "EARNY") ~ "Earnings",
      str_detect(VAR, "_HW")   ~ "Annual hours",
      str_detect(VAR, "PERM")  ~ "Indefinite contract",
      TRUE ~ NA_character_
    ),
    
    # Put age groups in logical order
    age_group = factor(
      age_group,
      levels = c("25-34", "35-44", "45-54", "55-64"),
      ordered = TRUE
    ),
    
    # Put degree categories in logical order
    degree = factor(
      degree,
      levels = c("Master", "Doctorate")
    ),
    
    # Flag cases where more than one normal value was averaged
    duplicate_normal = n_values > 1
  ) %>%
  
  select(
    country_code = REF_AREA,
    country = `Reference area`,
    degree,
    age_group,
    outcome,
    value = value_mean,
    duplicate_normal,
    n_normal_values = n_values,
    value_min,
    value_max,
    difference,
    pct_difference,
    VAR
  ) %>%
  
  arrange(
    country,
    degree,
    outcome,
    age_group
  )

#Check that all variables got converted and no NA values exist
age_tidy %>%
  filter(
    is.na(degree) |
    is.na(age_group) |
    is.na(outcome)
  )

#Inspect coverage
age_tidy %>%
  count(degree, outcome, age_group)

#Check that no duplicate values remain
age_tidy %>%
  count(country, degree, age_group, outcome) %>%
  filter(n > 1)
```

Idea: an interactive table displaying numbers per group.

```{r}
age_tidy
```


PLOTTING

Plot 1
How do earnings, working hours, and job security differ across age groups, and are the patterns different for Master’s and Doctorate holders?

```{r}
#Only use consistent data
plot1_data <- age_tidy %>%
  filter(
    !is.na(country),
    !is.na(degree),
    !is.na(age_group),
    !is.na(outcome),
    !is.na(value)
  ) %>%
  
  group_by(country, degree, outcome) %>%
  
  # Keep only cases where all 4 age groups are available
  filter(n_distinct(age_group) == 4) %>%
  
  ungroup()

#Number of countries:
plot1_data %>%
  distinct(country, degree, outcome) %>%
  count(degree, outcome)

#Create age-dependent change relative to 25-34 years
plot1_data <- plot1_data %>%
  group_by(country, degree, outcome) %>%
  mutate(
    baseline_25_34 = value[age_group == "25-34"],
    
    age_change = case_when(
      
      # Percentage change
      outcome == "Earnings" ~
        100 * (value / baseline_25_34 - 1),
      
      outcome == "Annual hours" ~
        100 * (value / baseline_25_34 - 1),
      
      outcome == "Indefinite contract" ~
        100 * (value / baseline_25_34 - 1)
    )
  ) %>%
  ungroup()

#Check
plot1_data %>%
  select(
    country,
    degree,
    outcome,
    age_group,
    value,
    baseline_25_34,
    age_change
  ) %>%
  arrange(country, degree, outcome, age_group)

#Summarize across countries, use median and 25th and 75th percentiles
plot1_summary <- plot1_data %>%
  group_by(degree, outcome, age_group) %>%
  summarise(
    median_change = median(age_change, na.rm = TRUE),
    q25 = quantile(age_change, 0.25, na.rm = TRUE),
    q75 = quantile(age_change, 0.75, na.rm = TRUE),
    n_countries = n(),
    .groups = "drop"
  )

#inspect
plot1_summary

#Prepare facets
plot1_summary <- plot1_summary %>%
  mutate(
    outcome_label = case_when(
      outcome == "Earnings" ~
        "Earnings",
      
      outcome == "Annual hours" ~
        "Annual working hours",
      
      outcome == "Indefinite contract" ~
        "Indefinite contracts"
    )
  )

#Plot
library(ggplot2)

ggplot(
  plot1_summary,
  aes(
    x = age_group,
    y = median_change,
    group = degree
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.4
  ) +
  
  geom_line(
    aes(linetype = degree),
    linewidth = 1
  ) +
  
  geom_point(
    aes(shape = degree),
    size = 3
  ) +
  
  facet_wrap(
    ~ outcome_label,
    scales = "free_y"
  ) +
  
  labs(
    title = "Labour-market outcomes in age groups",
    subtitle = "Median within-country change relative to ages 25-34",
    x = "Age group",
    y = "Change relative to ages 25-34 (%)",
    linetype = "Degree",
    shape = "Degree",
    caption = "Only country-degree groups containing all four age groups displayed."
  ) +
  
  theme_minimal() +
  
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

```

What can we learn from here?
- Working hours do not change much during the course of the career, except for master´s degree holders, who tend to work 4% less at later ages (55-64 years).
- Doctorate degree holders are more likely to be hired with indefinite contracts at the later stages of their career.
- Earnings differ only in the later stages of life - from 45-54 years higher pay is evident for doctorate degree holders. 


PLOT2
Does the age–earnings pattern depend on the country?

```{r}
#Find countries with 4 age groups for both MSc and PhD
balanced_countries <- age_tidy %>%
  filter(
    outcome == "Earnings",
    !is.na(value)
  ) %>%
  
  group_by(country, degree) %>%
  summarise(
    n_age_groups = n_distinct(age_group),
    .groups = "drop"
  ) %>%
  
  # Complete age profile for the degree
  filter(n_age_groups == 4) %>%
  
  # Now check whether BOTH degrees are complete
  count(country, name = "n_degrees") %>%
  filter(n_degrees == 2) %>%
  
  pull(country)


#Look, what remained
balanced_countries
length(balanced_countries)
print(paste0("Only ", length(balanced_countries), 
" remain, including ", paste(balanced_countries, collapse = ", "),"."))

#Organize data for heatmap
heatmap_data <- age_tidy %>%
  filter(
    outcome == "Earnings",
    country %in% balanced_countries
  ) %>%
  
  group_by(country, degree) %>%
  
  mutate(
    earnings_25_34 = value[age_group == "25-34"],
    
    earnings_change =
      100 * (value / earnings_25_34 - 1)
  ) %>%
  
  ungroup()

#Sort countries according to their older age earnings
country_order <- heatmap_data %>%
  filter(age_group == "55-64") %>%
  group_by(country) %>%
  summarise(
    senior_premium = mean(earnings_change, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(senior_premium) %>%
  pull(country)

#Apply ordering
heatmap_data <- heatmap_data %>%
  mutate(
    country = factor(
      country,
      levels = country_order
    )
  )

#Heatmap
ggplot(
  heatmap_data,
  aes(
    x = age_group,
    y = country,
    fill = earnings_change
  )
) +
  
  geom_tile(
    linewidth = 0.5
  ) +
  
  geom_text(
    aes(label = round(earnings_change, 0)),
    size = 3
  ) +
  
  facet_wrap(~ degree) +
  
  scale_fill_gradient2(
    midpoint = 0,
    name = "% difference\nfrom 25-34"
  ) +
  
  labs(
    title = "The age-earnings gradient across countries",
    subtitle = "Earnings in each age group relative to ages 25-34 within the same country and degree",
    x = "Age group",
    y = NULL,
    caption = "Balanced sample: countries with all four age groups for both Master and Doctorate holders."
  ) +
  
  theme_minimal() +
  
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )


```

What do the results tell us?
Aging is not generally a bad thing - in all countries, people in higher age groups earn more than young people between 25-34. You can check out your home country from plot X.

Plot 3
Is it worth to have a PhD over MSc at older age?

```{r}
#Keep youngst and oldest groups
age_premium <- age_tidy %>%
  filter(
    outcome == "Earnings",
    age_group %in% c("25-34", "55-64"),
    !is.na(value)
  ) %>%
  
  select(
    country,
    degree,
    age_group,
    value
  )


#Make a wide table
age_premium <- age_premium %>%
  pivot_wider(
    names_from = age_group,
    values_from = value
  )

#Require both age groups
age_premium <- age_premium %>%
  drop_na(`25-34`, `55-64`)

#Calculate age earnings premium
age_premium <- age_premium %>%
  mutate(
    age_premium =
      100 * (`55-64` / `25-34` - 1)
  )

#Reformat table to compare PhD vs MSc
degree_comparison <- age_premium %>%
  select(
    country,
    degree,
    age_premium
  ) %>%
  
  pivot_wider(
    names_from = degree,
    values_from = age_premium
  ) %>%
    drop_na(Master, Doctorate) # We need both degrees to compare them

#PhD - MSc
degree_comparison <- degree_comparison %>%
  mutate(
    phd_minus_master =
      Doctorate - Master
  ) %>%
  
  arrange(phd_minus_master)

#Plot
ggplot(
  degree_comparison,
  aes(
    x = reorder(country, phd_minus_master),
    y = phd_minus_master
  )
) +
  
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  
  geom_segment(
    aes(
      xend = reorder(country, phd_minus_master),
      y = 0,
      yend = phd_minus_master
    ),
    linewidth = 0.7
  ) +
  
  geom_point(
    size = 3
  ) +
  
  coord_flip() +
  
  labs(
    title = "Is the age-earnings gradient stronger for doctorate holders?",
    subtitle = "Difference between Doctorate and Master's 55-64 vs 25-34 earnings premiums",
    x = NULL,
    y = "Doctorate minus Master's age premium (percentage points)",
    caption = "Positive values indicate a stronger age-earnings gradient among doctorate holders."
  ) +
  
  theme_minimal()


#Add values to plot
ggplot(
  degree_comparison,
  aes(
    x = reorder(country, phd_minus_master),
    y = phd_minus_master
  )
) +
  
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  
  geom_segment(
    aes(
      xend = reorder(country, phd_minus_master),
      y = 0,
      yend = phd_minus_master
    ),
    linewidth = 0.7
  ) +
  
  geom_point(
    size = 3
  ) +
  
  geom_text(
    aes(
      label = paste0(
        round(phd_minus_master, 1),
        " pp"
      )
    ),
    hjust = ifelse(
      degree_comparison$phd_minus_master >= 0,
      -0.2,
      1.2
    ),
    size = 3.2
  ) +
  
  coord_flip() +
  
  labs(
    title = "Is the age-earnings gradient stronger for doctorate holders?",
    subtitle = "Doctorate age premium minus Master's age premium",
    x = NULL,
    y = "Difference in age premium (percentage points)",
    caption = "Age premium = percentage difference in earnings between ages 55-64 and 25-34."
  ) +
  
  theme_minimal()





```


```{r}
#| label: setup
#| message: false
#| cache: true
library(dplyr)
library(ggplot2)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale

```


```{r}
#| label: print-time
#| message: false
cat(format(Sys.time()))
```