---
title: "Happiness on a map"
author: Raphaël Mauron
date: today
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

Here, we can use an interactive map to visualize the input data and some of the results.

## Load, explore and prepare the data

First, we need to load and make sure of what we have as data.

```{r}
#| label: setup
#| message: false
#| cache: true

library(dplyr)
library(ggplot2)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale
library(readr)
library(giscoR)
library(sf)
library(leaflet)
library(patchwork)
library(DT)
library(tidyverse)
```


```{r}
#| label: load data
#| message: false
data_table <- readr::read_csv("./filtered_reICO_data.csv")
```


### some exploration

```{r}
#| label: checkdata
#| message: false
print(glimpse(data_table))
print(head(unique(data_table$Indicator), n = 15))
print(paste0("and ", length(unique(data_table$Indicator))-15,  " more..."))
```

All the necessary information are located in the `data_table$Indicator` column. We need to fetch some more information from there.

### Pivot wider and reformat
So education lvl, age groups, different percentages, working hours, etc. become separate columns.


```{r}
#| label: wrangle
#| message: false

# VAR is a structured code: "<numerator>X<denominator>" for shares, or a plain
# code for levels. The numerator encodes education, measure and age band, e.g.
#   PHD_TOC_PERM_35_44XPHD_TOC_35_44   MAS_EARNY_25_34   TOTWB_HAPPY_GESMTXTOTWB
measure_labels <- c(
  "_TOC_PERM"        = "pct_permanent_contract",
  "_HW"              = "hours_worked",
  "_EARNY"           = "earnings_gross_yearly",
  "_HAPPY_GESMT"     = "pct_happy",
  "_FINS_MEDHIGH"    = "pct_satisfied_finances",
  "_TIMEUSE_MEDHIGH" = "pct_satisfied_leisure"
)

# Observed values first, then caveated, then imputed. Q is always a missing value.
status_priority <- c("A", "B", "D", "U", "V", "I", "Q")

data_long <- data_table |>
  mutate(
    num_code  = str_remove(VAR, "X.*$"),
    core_code = num_code |>
      str_remove("^(TOT|MAS|PHD)(WB)?") |>
      str_remove("_\\d{2}_\\d{2}$"),
    education = factor(str_extract(num_code, "^(TOT|MAS|PHD)"),
                       levels = c("TOT", "MAS", "PHD"),
                       labels = c("Total population", "Master's", "Doctorate")),
    age_group = num_code |>
      str_extract("\\d{2}_\\d{2}") |>
      str_replace("_", "-") |>
      replace_na("All ages") |>
      factor(levels = c("All ages", "25-34", "35-44", "45-54", "55-64")),
    measure   = unname(measure_labels[core_code])
  )

# Fail loud if a VAR code ever stops matching the pattern above.
stopifnot(!anyNA(data_long$education), !anyNA(data_long$age_group),
          !anyNA(data_long$measure))

# 154 country x VAR pairs appear 2-3 times under different observation statuses.
# Keep the best-status rows, then average whatever is still tied.
data_long <- data_long |>
  mutate(status_rank = match(OBS_STATUS_1, status_priority)) |>
  slice_min(status_rank, by = c(REF_AREA, VAR), with_ties = TRUE) |>
  summarise(
    value    = if (all(is.na(OBS_VALUE))) NA_real_ else mean(OBS_VALUE, na.rm = TRUE),
    n_source = n(),
    .by = c(REF_AREA, `Reference area`, VAR, Indicator,
            education, age_group, measure, `Observation status`)
  ) |>
  rename(iso3 = REF_AREA, country = `Reference area`, status = `Observation status`) |>
  arrange(country, education, age_group, measure)

# One row per country x education x age band, one column per measure.
data_wide <- data_long |>
  select(iso3, country, education, age_group, measure, value) |>
  pivot_wider(names_from = measure, values_from = value) |>
  arrange(country, education, age_group)

# Correct the country name for Turkey since was encoded as latin1, and now back to UTF-8
data_wide <- data_wide |> mutate(country = if_else(iso3 == "TUR", "Türkiye", country))

# The original Indicator text cannot be a column in the wide table (each row
# spans up to 6 indicators), so keep it as a lookup for axis labels / tooltips.
indicator_dict <- data_long |>
  distinct(education, age_group, measure, VAR, Indicator)
```


### Check data_wide briefly

```{r}
#| label: checkdata2
#| message: false
print(head(data_wide))
print(unique(data_wide$education))
print(unique(data_wide$age_group))
print(range(data_wide$pct_happy, na.rm = TRUE))
print(unique(data_wide$country))
```


### Save data_wide as new csv

```{r}
#| label: save data wide
#| message: false
write_csv(data_wide, file = "./data_wide.csv")
```


### Load the new formated data

```{r}
#| label: load data wide
#| message: false
data_wide <- read_csv("./data_wide.csv")
glimpse(data_wide)
```


## Add geometry/maps borders

To prepare the plotting into maps, we need some more information, notably the 
borders of the country present.

`giscoR` serves the official Eurostat GISCO boundaries


```{r}
#| label: geometry
#| cache: true

borders <- giscoR::gisco_get_countries( #get coordinates of the country present in our data_wide$iso3
  resolution = "20",              # 1:20 million — plenty for a country map
  epsg       = "4326",
  country    = data_wide$iso3
)

borders <- sf::st_make_valid(borders) #check whether the borders are valid for sf

# France, Spain, Portugal, Norway carry overseas/Arctic territory in GISCO,
# so never trust the automatic bounds — set them explicitly.
eu_bbox <- c(xmin = -12, ymin = 34, xmax = 32, ymax = 71)

head(borders)
```

Borders is a tibble with each row representing a single country's borders information.

### Join data_wide and borders

Since for the plotting we want each country to appear only exactly once, we have to
select 

```{r}
#| label: map-data

# The slice the maps are fixed at. "Doctorate"/"Master's" x "All ages" are the
# only combinations that carry all six measures (see the coverage note above).
map_education <- "Master's"
map_age_group <- "All ages"

# `borders` stays geometry-only: one row per country, no measures attached.
# data_wide has 11 rows per country (education x age band), so the join has to
# happen against one slice at a time — see the map chunk.
eu_sf <- borders |>
  left_join(filter(data_wide, education == map_education, age_group == map_age_group),
            by = c("ISO3_CODE" = "iso3"))

stopifnot(nrow(eu_sf) == nrow(borders))   # one row, one geometry, one country

metrics <- tribble(
  ~col,                     ~label,                     ~palette,
  "pct_happy",              "Happy (%)",                "YlGnBu",
  "pct_satisfied_finances", "Satisfied, finances (%)",  "YlGnBu",
  "pct_satisfied_leisure",  "Satisfied, leisure (%)",   "YlGnBu",
  "pct_permanent_contract", "Permanent contract (%)",   "YlGnBu",
  "earnings_gross_yearly",  "Gross yearly earnings",    "YlOrBr",
  "hours_worked",           "Hours worked per year",    "YlOrBr"
)

# Fail loud if a slice ever stops carrying one of the layers.
stopifnot(all(metrics$col %in% names(eu_sf)),
          !sapply(metrics$col, \(c) all(is.na(eu_sf[[c]]))))
```



### All six metrics at once: a 2 x 3 panel

`facet_wrap()` is the wrong tool here. It shares one fill scale across panels,
and these six measures live on incompatible ranges — percentages in the tens,
hours in the thousands, earnings in the tens of thousands. A shared scale would
render four panels uniformly pale and one uniformly dark. Six separate plots
stitched with `patchwork` give each panel a scale, and a legend, of its own.

`coord_sf()` does the cropping: Brazil, Colombia, Korea and the USA are in
`data_wide` and would otherwise show up in every panel.

```{r}
#| label: metric-grid
#| fig-width: 12
#| fig-height: 8
map_panel <- function(col, label, palette) {
  ggplot(eu_sf) +
    geom_sf(aes(fill = .data[[col]]), colour = "white", linewidth = 0.15) +
    coord_sf(xlim = eu_bbox[c("xmin", "xmax")],
             ylim = eu_bbox[c("ymin", "ymax")], expand = FALSE) +
    # direction = 1 because distiller defaults to -1, which would paint low
    # values dark and silently invert every panel against the leaflet map.
    scale_fill_distiller(palette = palette, direction = 1, na.value = "grey90",
                         labels = scales::label_number(big.mark = " ")) +
    labs(title = label, fill = NULL) +
    theme_void(base_size = 9) +
    theme(
      plot.title        = element_text(size = 10, hjust = 0, margin = margin(b = 4)),
      # without the bottom margin the row-1 colourbars land on the row-2 titles
      # and the row-2 colourbars get clipped by the figure edge.
      plot.margin       = margin(4, 6, 14, 6),
      legend.position   = "bottom",
      legend.margin     = margin(t = 6, b = 0),
      legend.text       = element_text(size = 7),
      legend.key.height = unit(3, "pt"),
      legend.key.width  = unit(28, "pt")
    )
}

wrap_plots(pmap(metrics, map_panel), nrow = 2, ncol = 3) +
  plot_annotation(title = paste(map_education, "-", map_age_group))
```

## Correct the income values of Croatia:

```{r}
# Croatia reports earnings in kuna while every other country is on the survey's
# harmonised unit. Mean gross yearly earnings across all EARNY variables:
# HRV 322 743, against 122 411 for Luxembourg, the next highest -- a 2.6x gap
# with no counterpart in any other indicator. Dividing by the irrevocable
# HRK/EUR rate fixed at Croatia's euro entry (1 Jan 2023) lands Croatia at
# 42 835, between Estonia and Poland, which matches its position elsewhere.
HRK_PER_EUR <- 7.53450

# Before/after, for the write-up. Run this *before* the mutate below.
data_wide |>
  filter(iso3 == "HRV", !is.na(earnings_gross_yearly)) |>
  transmute(education, age_group,
            as_reported = round(earnings_gross_yearly),
            in_euro     = round(earnings_gross_yearly / HRK_PER_EUR))

data_wide <- data_wide |>
  mutate(earnings_gross_yearly = if_else(iso3 == "HRV",
                                         earnings_gross_yearly / HRK_PER_EUR,
                                         earnings_gross_yearly))

```


## The slice both maps draw

```{r}
#| label: map-data2

# The slice the maps are fixed at. "Doctorate"/"Master's" x "All ages" are the
# only combinations that carry all six measures (see the coverage note above).
map_education <- "Master's"
map_age_group <- "All ages"

eu_sf <- borders |>
  left_join(filter(data_wide, education == map_education, age_group == map_age_group),
            by = c("ISO3_CODE" = "iso3"))

stopifnot(nrow(eu_sf) == nrow(borders))   # one row, one geometry, one country

metrics <- tribble(
  ~col,                     ~label,                     ~palette,
  "pct_happy",              "Happy (%)",                "YlGnBu",
  "pct_satisfied_finances", "Satisfied, finances (%)",  "YlGnBu",
  "pct_satisfied_leisure",  "Satisfied, leisure (%)",   "YlGnBu",
  "pct_permanent_contract", "Permanent contract (%)",   "YlGnBu",
  "earnings_gross_yearly",  "Gross yearly earnings",    "YlOrBr",
  "hours_worked",           "Hours worked per year",    "YlOrBr"
)

# Fail loud if a slice ever stops carrying one of the layers.
stopifnot(all(metrics$col %in% names(eu_sf)),
          !sapply(metrics$col, \(c) all(is.na(eu_sf[[c]]))))
```


## All six metrics at once: a 2 x 3 panel

`facet_wrap()` is the wrong tool here. It shares one fill scale across panels,
and these six measures live on incompatible ranges — percentages in the tens,
hours in the thousands, earnings in the tens of thousands. A shared scale would
render four panels uniformly pale and one uniformly dark. Six separate plots
stitched with `patchwork` give each panel a scale, and a legend, of its own.

`coord_sf()` does the cropping: Brazil, Colombia, Korea and the USA are in
`data_wide` and would otherwise show up in every panel.

```{r}
#| label: metric-grid2
#| fig-width: 12
#| fig-height: 8
map_panel <- function(col, label, palette) {
  ggplot(eu_sf) +
    geom_sf(aes(fill = .data[[col]]), colour = "white", linewidth = 0.15) +
    coord_sf(xlim = eu_bbox[c("xmin", "xmax")],
             ylim = eu_bbox[c("ymin", "ymax")], expand = FALSE) +
    # direction = 1 because distiller defaults to -1, which would paint low
    # values dark and silently invert every panel against the leaflet map.
    scale_fill_distiller(palette = palette, direction = 1, na.value = "grey90",
                         labels = scales::label_number(big.mark = " ")) +
    labs(title = label, fill = NULL) +
    theme_void(base_size = 9) +
    theme(
      plot.title        = element_text(size = 10, hjust = 0, margin = margin(b = 4)),
      # without the bottom margin the row-1 colourbars land on the row-2 titles
      # and the row-2 colourbars get clipped by the figure edge.
      plot.margin       = margin(4, 6, 14, 6),
      legend.position   = "bottom",
      legend.margin     = margin(t = 6, b = 0),
      legend.text       = element_text(size = 7),
      legend.key.height = unit(3, "pt"),
      legend.key.width  = unit(28, "pt")
    )
}

wrap_plots(pmap(metrics, map_panel), nrow = 2, ncol = 3) +
  plot_annotation(title = paste(map_education, "-", map_age_group))
```


## With correct Croatia currency


**This is the recommended default.** Leaflet is the most mature interactive
mapping package in R, it renders `sf` natively, and `addLayersControl()` gives
you exactly the radio-button toggle you described — no Shiny server needed.

The pattern: add one `addPolygons()` layer per metric, each with a `group =`
label, then wire the group names into `addLayersControl()`. Legends accept the
same `group =` argument, so they appear and disappear with their layer.

The radio buttons can only carry one dimension, and one geometry can only carry
one row, so education and age band are fixed and the six measures are the
toggle. Coverage forces that choice: the wellbeing measures are reported only at
`All ages`, the labour measures only for Master's and Doctorate. `Master's` and
`Doctorate` at `All ages` are therefore the only two slices with all six.


```{r}
#| label: leaflet-map2

add_metric_layer <- function(map, col, label, palette) {
  v   <- eu_sf[[col]]
  pal <- colorNumeric(palette, domain = v, na.color = "#e8e8e8")
  # 4-6 countries are missing per measure; say so rather than printing "NA".
  txt <- ifelse(is.na(v), "no data",
                formatC(v, format = "f", digits = 1, big.mark = " "))

  map |>
    addPolygons(
      data         = eu_sf,
      group        = label,
      fillColor    = ~ pal(v),
      fillOpacity  = 0.8,
      color        = "white",
      weight       = 1,
      label        = ~ sprintf("%s: %s", NAME_ENGL, txt) |> lapply(htmltools::HTML),
      highlightOptions = highlightOptions(weight = 3, color = "#333", bringToFront = TRUE)
    ) |>
    addLegend(
      position = "bottomright", pal = pal, values = v,
      title = label, group = label, opacity = 0.9
    )
}

m <- leaflet(options = leafletOptions(minZoom = 3)) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>
  fitBounds(eu_bbox[["xmin"]], eu_bbox[["ymin"]], eu_bbox[["xmax"]], eu_bbox[["ymax"]])

m <- reduce(
  seq_len(nrow(metrics)),
  \(map, i) add_metric_layer(map, metrics$col[i], metrics$label[i], metrics$palette[i]),
  .init = m
)

m |>
  addLayersControl(
    baseGroups    = metrics$label,           # radio buttons: one metric at a time
    options       = layersControlOptions(collapsed = FALSE)
  ) |>
  # leaflet ties legend visibility to overlayadd/overlayremove/groupadd/
  # groupremove and has no baselayerchange handler, so with baseGroups every
  # legend is added at init and never removed -- all six stack up at once.
  # Translate the event the radio buttons do fire into the ones it listens for.
  htmlwidgets::onRender("
function(el, x) {
  var map = this;
  var groups = x.calls
    .filter(function(c) { return c.method === 'addLegend'; })
    .map(function(c) { return c.args[0].group; });

  // addLayersControl already leaves only the first polygon layer visible;
  // do the same for the legends.
  groups.slice(1).forEach(function(g) { map.fire('groupremove', {name: g}); });

  var shown = groups[0];
  map.on('baselayerchange', function(e) {
    if (e.name === shown) return;
    map.fire('groupremove', {name: shown});
    map.fire('groupadd',    {name: e.name});
    shown = e.name;
  });
}")
```


```{r}
datatable(data_wide)
```


```{r}

```

```{r}

```


```{r}

```


```{r}

```

```{r}

```


```{r}

```


```{r}

```



```{r}
#| label: print-time
#| message: false
sessionInfo()
```