---
title: "Happiness on a map"
author: Raphaël Mauron
date: today
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

Here, we can use an interactive map to visualize the input data and some of the results.

## Load the data

First, we need to load and make sure of what we have as data.

```{r}
#| label: setup
#| message: false

# Not cached: on a cache hit knitr restores the chunk's *objects* but does not
# re-attach its packages, so every later chunk loses str_remove(), glimpse(),
# pivot_wider() and friends. Attaching them takes a second anyway.
library(dplyr)
library(stringr)
library(ggplot2)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale
library(readr)
library(giscoR)
library(sf)
library(leaflet)
library(patchwork)
library(DT)
library(tidyverse)
```



### Load the new formated data

```{r}
#| label: load-data-wide
#| message: false
data_wide <- read_csv("./data_wide.csv")
dplyr::glimpse(data_wide)
```


## Add geometry/maps borders

To prepare the plotting into maps, we need some more information, notably the 
borders of the country present.

`giscoR` serves the official Eurostat GISCO boundaries


```{r}
#| label: geometry
#| cache: true

# Cached, and nothing invalidates it automatically: `dependson` only tracks
# chunks that are themselves cached, and load-data-wide is not. If the country
# set in data_wide ever changes, delete analysis_4_cache/ by hand.

borders <- giscoR::gisco_get_countries( #get coordinates of the country present in our data_wide$iso3
  resolution = "20",              # 1:20 million — plenty for a country map
  epsg       = "4326",
  country    = data_wide$iso3
)

borders <- sf::st_make_valid(borders) #check whether the borders are valid for sf

# France, Spain, Portugal, Norway carry overseas/Arctic territory in GISCO,
# so never trust the automatic bounds — set them explicitly.
eu_bbox <- c(xmin = -12, ymin = 34, xmax = 32, ymax = 71)

head(borders)
```

Borders is a tibble with each row representing a single country's borders information.

### Join data_wide and borders

Since for the plotting we want each country to appear only exactly once, we have to
select 

```{r}
#| label: map-data

# The slice the maps are fixed at. "Doctorate"/"Master's" x "All ages" are the
# only combinations that carry all six measures (see the coverage note above).
map_education <- "Master's"
map_age_group <- "All ages"

# `borders` stays geometry-only: one row per country, no measures attached.
# data_wide has 11 rows per country (education x age band), so the join has to
# happen against one slice at a time — see the map chunk.
eu_sf <- borders |>
  left_join(filter(data_wide, education == map_education, age_group == map_age_group),
            by = c("ISO3_CODE" = "iso3"))

stopifnot(nrow(eu_sf) == nrow(borders))   # one row, one geometry, one country

metrics <- tribble(
  ~col,                     ~label,                     ~palette,
  "pct_happy",              "Happy (%)",                "YlGnBu",
  "pct_satisfied_finances", "Satisfied, finances (%)",  "YlGnBu",
  "pct_satisfied_leisure",  "Satisfied, leisure (%)",   "YlGnBu",
  "pct_permanent_contract", "Permanent contract (%)",   "YlGnBu",
  "earnings_gross_yearly",  "Gross yearly earnings",    "YlOrBr",
  "hours_worked",           "Hours worked per year",    "YlOrBr"
)

# Fail loud if a slice ever stops carrying one of the layers.
stopifnot(all(metrics$col %in% names(eu_sf)),
          !sapply(metrics$col, \(c) all(is.na(eu_sf[[c]]))))
```



### All six metrics at once: a 2 x 3 panel

`facet_wrap()` is the wrong tool here. It shares one fill scale across panels,
and these six measures live on incompatible ranges — percentages in the tens,
hours in the thousands, earnings in the tens of thousands. A shared scale would
render four panels uniformly pale and one uniformly dark. Six separate plots
stitched with `patchwork` give each panel a scale, and a legend, of its own.

`coord_sf()` does the cropping: Brazil, Colombia, Korea and the USA are in
`data_wide` and would otherwise show up in every panel.

```{r}
#| label: metric-grid
#| fig-width: 12
#| fig-height: 8
map_panel <- function(col, label, palette) {
  ggplot(eu_sf) +
    geom_sf(aes(fill = .data[[col]]), colour = "white", linewidth = 0.15) +
    coord_sf(xlim = eu_bbox[c("xmin", "xmax")],
             ylim = eu_bbox[c("ymin", "ymax")], expand = FALSE) +
    # direction = 1 because distiller defaults to -1, which would paint low
    # values dark and silently invert every panel against the leaflet map.
    scale_fill_distiller(palette = palette, direction = 1, na.value = "grey90",
                         labels = scales::label_number(big.mark = " ")) +
    labs(title = label, fill = NULL) +
    theme_void(base_size = 9) +
    theme(
      plot.title        = element_text(size = 10, hjust = 0, margin = margin(b = 4)),
      # without the bottom margin the row-1 colourbars land on the row-2 titles
      # and the row-2 colourbars get clipped by the figure edge.
      plot.margin       = margin(4, 6, 14, 6),
      legend.position   = "bottom",
      legend.margin     = margin(t = 6, b = 0),
      legend.text       = element_text(size = 7),
      legend.key.height = unit(3, "pt"),
      legend.key.width  = unit(28, "pt")
    )
}

wrap_plots(pmap(metrics, map_panel), nrow = 2, ncol = 3) +
  plot_annotation(title = paste(map_education, "-", map_age_group))
```

## Correct the income values of Croatia:

```{r}
# Croatia reports earnings in kuna while every other country is on the survey's
# harmonised unit. Mean gross yearly earnings across all EARNY variables:
# HRV 322 743, against 122 411 for Luxembourg, the next highest -- a 2.6x gap
# with no counterpart in any other indicator. Dividing by the irrevocable
# HRK/EUR rate fixed at Croatia's euro entry (1 Jan 2023) lands Croatia at
# 42 835, between Estonia and Poland, which matches its position elsewhere.
HRK_PER_EUR <- 7.53450

# Before/after, for the write-up. Run this *before* the mutate below.
data_wide |>
  filter(iso3 == "HRV", !is.na(earnings_gross_yearly)) |>
  transmute(education, age_group,
            as_reported = round(earnings_gross_yearly),
            in_euro     = round(earnings_gross_yearly / HRK_PER_EUR))

data_wide <- data_wide |>
  dplyr::mutate(earnings_gross_yearly = if_else(iso3 == "HRV",
                                         earnings_gross_yearly / HRK_PER_EUR,
                                         earnings_gross_yearly))

```


## The slice both maps draw

```{r}
#| label: map-data2

# The slice the maps are fixed at. "Doctorate"/"Master's" x "All ages" are the
# only combinations that carry all six measures (see the coverage note above).
map_education <- "Master's"
map_age_group <- "All ages"

eu_sf <- borders |>
  left_join(filter(data_wide, education == map_education, age_group == map_age_group),
            by = c("ISO3_CODE" = "iso3"))

stopifnot(nrow(eu_sf) == nrow(borders))   # one row, one geometry, one country

metrics <- tribble(
  ~col,                     ~label,                     ~palette,
  "pct_happy",              "Happy (%)",                "YlGnBu",
  "pct_satisfied_finances", "Satisfied, finances (%)",  "YlGnBu",
  "pct_satisfied_leisure",  "Satisfied, leisure (%)",   "YlGnBu",
  "pct_permanent_contract", "Permanent contract (%)",   "YlGnBu",
  "earnings_gross_yearly",  "Gross yearly earnings",    "YlOrBr",
  "hours_worked",           "Hours worked per year",    "YlOrBr"
)

# Fail loud if a slice ever stops carrying one of the layers.
stopifnot(all(metrics$col %in% names(eu_sf)),
          !sapply(metrics$col, \(c) all(is.na(eu_sf[[c]]))))
```


## All six metrics at once: a 2 x 3 panel

`facet_wrap()` is the wrong tool here. It shares one fill scale across panels,
and these six measures live on incompatible ranges — percentages in the tens,
hours in the thousands, earnings in the tens of thousands. A shared scale would
render four panels uniformly pale and one uniformly dark. Six separate plots
stitched with `patchwork` give each panel a scale, and a legend, of its own.

`coord_sf()` does the cropping: Brazil, Colombia, Korea and the USA are in
`data_wide` and would otherwise show up in every panel.

```{r}
#| label: metric-grid2
#| fig-width: 12
#| fig-height: 8
map_panel <- function(col, label, palette) {
  ggplot(eu_sf) +
    geom_sf(aes(fill = .data[[col]]), colour = "white", linewidth = 0.15) +
    coord_sf(xlim = eu_bbox[c("xmin", "xmax")],
             ylim = eu_bbox[c("ymin", "ymax")], expand = FALSE) +
    # direction = 1 because distiller defaults to -1, which would paint low
    # values dark and silently invert every panel against the leaflet map.
    scale_fill_distiller(palette = palette, direction = 1, na.value = "grey90",
                         labels = scales::label_number(big.mark = " ")) +
    labs(title = label, fill = NULL) +
    theme_void(base_size = 9) +
    theme(
      plot.title        = element_text(size = 10, hjust = 0, margin = margin(b = 4)),
      # without the bottom margin the row-1 colourbars land on the row-2 titles
      # and the row-2 colourbars get clipped by the figure edge.
      plot.margin       = margin(4, 6, 14, 6),
      legend.position   = "bottom",
      legend.margin     = margin(t = 6, b = 0),
      legend.text       = element_text(size = 7),
      legend.key.height = unit(3, "pt"),
      legend.key.width  = unit(28, "pt")
    )
}

wrap_plots(pmap(metrics, map_panel), nrow = 2, ncol = 3) +
  plot_annotation(title = paste(map_education, "-", map_age_group))
```


## Master's over Doctorate: the same six metrics, twelve maps

The grid above fixes education, so it cannot show what the education level does to
each measure. Here the six metrics are the columns and the two education levels
are the rows: read down a column to compare Master's against Doctorate, across a
row to compare measures within one level.

Both cells of a column share one `limits`, and only with each other. Without
that, `scale_fill_distiller()` would fit an independent range per panel and the
darkest shade would mean something different in the top cell than in the bottom
one -- which is exactly the comparison the column is there to support. Across
columns the scales stay separate, for the reason given above: the six measures
are on incompatible ranges.

Each metric also carries its own gradient, so a column can be told from its
neighbours by hue alone. That trades away the cool/warm split of the grid above,
where the colour separated percentages from money and time.

```{r}
#| label: metric-grid-education
#| fig-width: 14
#| fig-height: 8
#| column: page

# One slice per education level, joined exactly as above so each still carries
# one geometry per country. `Master's` repeats the slice the grid above draws;
# it is rebuilt here rather than reusing `eu_sf` so this figure keeps its two
# levels even if `map_education` is later pointed somewhere else.
edu_slice <- function(edu) {
  s <- borders |>
    left_join(filter(data_wide, education == edu, age_group == map_age_group),
              by = c("ISO3_CODE" = "iso3"))
  stopifnot(nrow(s) == nrow(borders))   # one row, one geometry, one country
  s
}

sf_mas <- edu_slice("Master's")
sf_phd <- edu_slice("Doctorate")

# Fail loud if a metric ever goes empty in both slices: `range()` below would
# otherwise hand scale_fill_distiller() an infinite limit.
stopifnot(all(metrics$col %in% names(sf_phd)),
          !sapply(metrics$col, \(c) all(is.na(c(sf_mas[[c]], sf_phd[[c]])))))

# One gradient per metric. All six are ColorBrewer sequential ramps, so each is
# still colour-blind safe on its own. Purples and PuRd are the closest pair of
# the six, which is why they are not neighbouring columns; Greys is left out
# because grey90 below already means "no data". Local to this figure --
# `metrics$palette` still drives the 6-panel grid and the leaflet map.
edu_palettes <- c(
  pct_happy              = "PuRd",
  pct_satisfied_finances = "Blues",
  pct_satisfied_leisure  = "Greens",
  pct_permanent_contract = "Purples",
  earnings_gross_yearly  = "YlOrBr",
  hours_worked           = "OrRd"
)
stopifnot(setequal(names(edu_palettes), metrics$col))

# Per metric, not per panel: both rows of a column read off the same scale.
edu_limits <- setNames(
  lapply(metrics$col, \(col) range(c(sf_mas[[col]], sf_phd[[col]]), na.rm = TRUE)),
  metrics$col
)

# Same recipe as map_panel(), with the slice, the gradient and the fill range
# passed in, and the education level in a subtitle.
map_panel_edu <- function(sf_slice, col, palette, limits, title, subtitle) {
  ggplot(sf_slice) +
    geom_sf(aes(fill = .data[[col]]), colour = "white", linewidth = 0.15) +
    coord_sf(xlim = eu_bbox[c("xmin", "xmax")],
             ylim = eu_bbox[c("ymin", "ymax")], expand = FALSE) +
    scale_fill_distiller(palette = palette, direction = 1, limits = limits,
                         na.value = "grey90",
                         labels = scales::label_number(big.mark = " ")) +
    labs(title = title, subtitle = subtitle, fill = NULL) +
    theme_void(base_size = 8) +
    theme(
      # 6 columns leaves ~2.3 in of width per panel, so the metric name goes on
      # the title line at 8 pt and the education level on its own line under it.
      plot.title        = element_text(size = 8, hjust = 0, margin = margin(b = 1)),
      plot.subtitle     = element_text(size = 7, hjust = 0, colour = "grey30",
                                       margin = margin(b = 4)),
      plot.margin       = margin(4, 4, 12, 4),
      legend.position   = "bottom",
      legend.text       = element_text(size = 6),
      legend.key.height = unit(3, "pt"),
      legend.key.width  = unit(20, "pt")
    )
}

# One row per education level, metrics in the order `metrics` lists them, so
# column j is the same metric in both rows. wrap_plots() fills by row, so the
# two rows are simply concatenated. The colourbar under a column is drawn twice:
# it is the same scale both times, which is the point.
edu_row <- function(sf_slice, edu) {
  pmap(metrics, function(col, label, ...)    # gradient comes from edu_palettes
    map_panel_edu(sf_slice, col, edu_palettes[[col]], edu_limits[[col]],
                  label, edu))
}

wrap_plots(c(edu_row(sf_mas, "Master's"), edu_row(sf_phd, "Doctorate")),
           nrow = 2, ncol = 6) +
  plot_annotation(title = paste("Master's over Doctorate -", map_age_group))
```


## With correct Croatia currency


**This is the recommended default.** Leaflet is the most mature interactive
mapping package in R, it renders `sf` natively, and `addLayersControl()` gives
you exactly the radio-button toggle you described — no Shiny server needed.

The pattern: add one `addPolygons()` layer per metric, each with a `group =`
label, then wire the group names into `addLayersControl()`. Legends accept the
same `group =` argument, so they appear and disappear with their layer.

The radio buttons can only carry one dimension, and one geometry can only carry
one row, so education and age band are fixed and the six measures are the
toggle. Coverage forces that choice: the wellbeing measures are reported only at
`All ages`, the labour measures only for Master's and Doctorate. `Master's` and
`Doctorate` at `All ages` are therefore the only two slices with all six.


```{r}
#| label: leaflet-map2

add_metric_layer <- function(map, col, label, palette) {
  v   <- eu_sf[[col]]
  pal <- colorNumeric(palette, domain = v, na.color = "#e8e8e8")
  # 4-6 countries are missing per measure; say so rather than printing "NA".
  txt <- ifelse(is.na(v), "no data",
                formatC(v, format = "f", digits = 1, big.mark = " "))

  map |>
    addPolygons(
      data         = eu_sf,
      group        = label,
      fillColor    = ~ pal(v),
      fillOpacity  = 0.8,
      color        = "white",
      weight       = 1,
      label        = ~ sprintf("%s: %s", NAME_ENGL, txt) |> lapply(htmltools::HTML),
      highlightOptions = highlightOptions(weight = 3, color = "#333", bringToFront = TRUE)
    ) |>
    addLegend(
      position = "bottomright", pal = pal, values = v,
      title = label, group = label, opacity = 0.9
    )
}

m <- leaflet(options = leafletOptions(minZoom = 3)) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>
  fitBounds(eu_bbox[["xmin"]], eu_bbox[["ymin"]], eu_bbox[["xmax"]], eu_bbox[["ymax"]])

m <- reduce(
  seq_len(nrow(metrics)),
  \(map, i) add_metric_layer(map, metrics$col[i], metrics$label[i], metrics$palette[i]),
  .init = m
)

m |>
  addLayersControl(
    baseGroups    = metrics$label,           # radio buttons: one metric at a time
    options       = layersControlOptions(collapsed = FALSE)
  ) |>
  # leaflet ties legend visibility to overlayadd/overlayremove/groupadd/
  # groupremove and has no baselayerchange handler, so with baseGroups every
  # legend is added at init and never removed -- all six stack up at once.
  # Translate the event the radio buttons do fire into the ones it listens for.
  htmlwidgets::onRender("
function(el, x) {
  var map = this;
  var groups = x.calls
    .filter(function(c) { return c.method === 'addLegend'; })
    .map(function(c) { return c.args[0].group; });

  // addLayersControl already leaves only the first polygon layer visible;
  // do the same for the legends.
  groups.slice(1).forEach(function(g) { map.fire('groupremove', {name: g}); });

  var shown = groups[0];
  map.on('baselayerchange', function(e) {
    if (e.name === shown) return;
    map.fire('groupremove', {name: shown});
    map.fire('groupadd',    {name: e.name});
    shown = e.name;
  });
}")
```


## Two metrics side by side: a Shiny companion

The map above toggles one metric at a time, which is the most `addLayersControl()`
can do — radio buttons carry one dimension, and one geometry carries one row. The
question it cannot answer is whether two metrics move together: for that you need
both maps on screen at once, plus the scatter that pairs them country by country.

That needs a server, so it lives in [`shinyMap.R`](shinyMap.R) rather than in this
document. Keeping it out of the `.qmd` is deliberate: `quarto render` would have to
switch the whole page to `server: shiny` to host it, and this project renders as a
static website. The chunk below is `eval: false`, so the render never touches it.

The app draws the same `data_wide` slices and reuses the leaflet recipe from
`add_metric_layer()` above — `colorNumeric()` fill, `no data` labels for the 4-6
countries missing per measure, the same `CartoDB.PositronNoLabels` tiles and the
same fixed `eu_bbox`. What changes is that each metric gets its own map rather
than being one toggled layer among six, which is what lets two of them sit side
by side. Two metric selectors drive six panels: map 1, map 2 and their scatter
for Master's on the first row, the same three for Doctorate on the second.

Two things it does that the static figures cannot. Clicking a country on any map,
or a point in any scatter, rings it in all six panels at once, which is how you
tell whether a country sits high on one metric and low on the other. And the age
band is selectable, with the metric menus annotated per band — coverage is uneven
(the wellbeing measures exist only at `All ages`, the labour measures only per
band), so the app says which pairs exist instead of quietly drawing blanks.

Two smaller carry-overs from the figures above: each scatter reports its own
Pearson and Spearman correlation with the `n` it was computed from, and a toggle
puts both rows of a column on one shared colour scale — the `edu_limits` trick
from the twelve-map grid, without which the darkest shade means something
different in the Master's map than in the Doctorate one.

Note that `shinyMap.R` re-applies the Croatia kuna correction from above. It reads
`data_wide.csv`, and that file is written earlier in this document, before the fix.

```{r}
#| label: launch-shiny
#| eval: false

# From the project root. eval: false above -- this never runs during a render.
# This is the fast path while developing: real R, instant reload.
shiny::runApp("shinyMap.R")
```

### Getting it into this page: R compiled to WebAssembly

The chunk above needs a live R process, which a rendered `.html` does not have.
To make the app work *in this page* the R process has to move into the reader's
browser, and that is what `shinylive` does: it ships webR -- R itself compiled to
WebAssembly -- alongside wasm builds of every package the app loads. No server, no
`shinyapps.io` account, and the result is still a static site.

Every package this app needs is available as a wasm binary from
[`repo.r-wasm.org`](https://repo.r-wasm.org): `shiny`, `bslib`, `dplyr`, `readr`,
`sf`, `leaflet` and `plotly`.

One thing had to change to make this work. `giscoR` fetched the boundaries over
the network at startup, which a browser cannot do -- cross-origin requests to the
GISCO service are blocked, and webR has no raw sockets. So the geometry is baked
to a file once, here, from the very same `borders` object the static maps above
are drawn from:

```{r}
#| label: save-borders

# 2 decimal places is ~1 km at these latitudes, invisible at 1:20 million, and
# it takes the file from 156 KB to 137 KB. Only the two columns the app uses.
sf::st_write(
  dplyr::select(borders, ISO3_CODE, NAME_ENGL),
  "borders.geojson",
  driver        = "GeoJSON",
  delete_dsn    = TRUE,
  quiet         = TRUE,
  layer_options = "COORDINATE_PRECISION=2"
)
```

Building the bundle is not a chunk in this document, and that is not a stylistic
choice. Quarto expands the globs under `project: resources:` *before* it renders
anything, so a `shinylive_app/` created partway through a render is discovered too
late to be copied into `_site/`. What you get is a blank white frame: `index.html`
does make it across, because Quarto spots it in the `src` below, while the 72 MB
runtime sitting next to it does not.

So the export lives in [`export_shinylive.R`](export_shinylive.R) and runs as a
`pre-render` step, declared in `_quarto.yml` alongside the `resources:` entry it
has to precede. It downloads webR and a wasm binary for every package the app
loads, so the first build takes minutes; `shinylive` caches them afterwards.
Pre-render scripts are not covered by `freeze: auto` and would otherwise re-export
on every single render, so the script skips the work unless `shinyMap.R` is newer
than the bundle. Delete `shinylive_app/` to force a rebuild after a data change.

```{r}
#| label: check-shinylive-export
#| echo: false

# The runtime, not just the page: index.html on its own renders as a blank frame,
# which is easy to mistake for a broken app rather than a missing export.
if (!dir.exists("shinylive_app/shinylive")) {
  warning("shinylive_app/ is missing or incomplete -- check that the ",
          "pre-render step in _quarto.yml ran; the frame below will be blank.",
          call. = FALSE)
}
```

::: {.column-screen-inset}
```{=html}
<iframe src="shinylive_app/index.html" width="100%" height="1040"
        style="border: 1px solid #dee2e6; border-radius: 4px;">
</iframe>
```
:::

Two caveats worth knowing. The first load is heavy -- webR plus the `sf` stack
(GDAL, PROJ and GEOS, all compiled to wasm) is tens of megabytes, so expect a
visible startup pause; it caches afterwards. And the wasm app does not see
`_brand.yml`, so it renders in the default Bootstrap theme rather than the site's
teal and Albert Sans, while the locally launched version picks the brand up.



```{r}
#| label: print-time
#| message: false
sessionInfo()
```